---
title: "Bike Case Study"
subtitle: 'Source: Google DAC Course'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(tidyverse)
library(readxl)
library(kableExtra)

# folder
folder_input <- '3_input'


```

### 1. Task

#### Bike-share Case Study

Cyclistic launched their bike-share service in 2016. Since then, the program has grown to a fleet of 5,824 bicycles that are geo-tracked and locked into a network of 692 stations across Chicago. The bike can be unlocked from one station and returned to any other station in the system anytime.

Cyclistic offers flexible pricing plans including single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchases annual memberships are Cyclistic members.

#### Task

The company wants to design a marketing strategy that aimed at converting casual riders into annual members. In order to do that, they want to understand:

-   How do annual members and casual riders use Cyclistic bikes differently?

-   Why would casual riders buy Cyclistic annual memberships?

-   How can Cyclistic use digital media to influence casual riders to become members?

This report is to answer how annual members and casual riders differ.

### 2. Data Description

Data source: Google Data Analytics Certificate on Cousera

#### Trip Data

Before 2020, the trip data was stored using the following metadata:

```{r metadata tripb20, echo=FALSE}

# import metadata trip before 2020
meta_tripb20 <- 
  read_xlsx(file.path(folder_input, 
                      'metadata_trip_b2020.xlsx'))

# show table
kable(meta_tripb20)


```

**Notes:**

-   Trips that did not include a start or end date are excluded

-   Trips less than 1 minute in duration are excluded

-   Trips greater than 24 hours in duration are excluded

-   Gender and birthday are only available for Subscribers

Since 2020, the data structure changed to:

```{r metadata trips20, echo=FALSE}

# import metadata trip since 2020
meta_trips20 <- 
  read_xlsx(file.path(folder_input,
                      'metadata_trip_s2020.xlsx'))

# show table
kable(meta_trips20)

```

#### Station Data

Store all the data of stations in the system

```{r metadata station, echo=FALSE}

# import meta data station
meta_station <- 
  read_xlsx(file.path(folder_input,
                      'metadata_station.xlsx'))

# show table
kable(meta_station)

```

### 3. Data Cleaning & Manipulation
